### 2025-06-14 lecturenotes_aima104/ch3_aima104.tex
chap3_aima104

  \begin{theorem}
    Zero is an eigenvalue of $A$ if and only if $A$ is not invertible.
  \end{theorem}
\section{Diagonalization}
\begin{mdframed}
  A square matrix A is said to be diagonalizable if, there exists an invertible matrix $P$ and a diagonal matrix $D$, such that $$A = PDP^{-1}.$$ If $A = PDP^{-1}$ then $A$ is said to be similar to $D$.
\end{mdframed}

\subsection{Similarity}
If $A$ an $B$ are $n \times n$ matrices, then $A$ \textbf{similar to} $B$ if there is an invertible matrix $P$ such that $$A = PBP^{-1}$$ or equivalently $B=P^{-1}AP$. Writing $Q=P^{-1}$, we can say that $B$ is similar to $A$, and we simply say $A$ and $B$ are similar.
\begin{theorem}
  If $n \times n$ matrices $A$ and $B$ are similar, then they have the same characteristic polynomial and hence the same eigenvalues (with the same multiplicities).
\end{theorem}
\begin{proof}
 \hspace{1cm} Given, $B=P^{-1}AP$ we have, $(B-\lambda I) = P^{-1}AP - \lambda P^{-1}P = P^{-1}(A- \lambda I) P$. Then
    \begin{align*}
      det(B-\lambda I) &= \det[P^{-1}(A- \lambda I) P] \\
                       &= \det(P^{-1}).\, \det(A- \lambda I).\, \det(P) \\
                       &= \det(P^{-1}).\, \det(P). \, \det(A- \lambda I) \\
                       &= \det(A- \lambda I)
    \end{align*}
  As, $\det(P^{-1}).\, \det(P) = \det(P^{-1}.P)=det(I) = 1$. Hence $A$ and $B$ has same characteristic polynomial. And the proof follows.
\end{proof}
\subsection{The Diagonalization Theorems}
Suppose $A$ is $n \times n$ matrix with $n$ different eigenvalues: $\lambda_1, \lambda_2, ..., \lambda_n$ so that the set of $n$ corresponding eigenvectors $v_1, v_2, ..., v_n$ are linearly independent. Now, $Av_1=\lambda _n v_1, Av_2=\lambda _2 v_2, ...., Av_n = \lambda _n v_n$. Let $P$ be the matrix whose columns are the eigenvectors, i.e $P = [v_1, \; v_2, \; ... \; v_n] $. Then
\begin{equation}
  \label{eq:1}
AP = [\lambda _1 v_1, \; \lambda_2 v_2, \; ... \; \lambda_n v_n] =
 [v_1, \; v_2, \; , ..., \; v_n] \begin{bmatrix}
  \lambda_1 & 0 & ... & 0 \\
  0 & \lambda_2 & ... & 0 \\
  ... \\
  0 & 0 & ... & \lambda_1 \\
\end{bmatrix} = PD
\end{equation}
where $ D = \begin{bmatrix}
  \lambda_1 & 0 & ... & 0 \\
  0 & \lambda_2 & ... & 0 \\
  ... \\
  0 & 0 & ... & \lambda_1 \\
\end{bmatrix} $, so that $AP = PD$ gives, $A = P^{-1} D P$. This is summarized as the following theorem.

@@

\subsection{Exercise}
Diagonalizable the following matrix, if possible.
$$A=
\begin{bmatrix}
  5 & 0 & 0 & 0 \\
  0 & 5 & 0 & 0 \\
  1 & 4 & -3 & 0 \\
  -1 & -2 & 0 & -3
\end{bmatrix}
$$
\end{document}

>>>>

  \begin{theorem}
    Zero is an eigenvalue of $A$ if and only if $A$ is not invertible.
  \end{theorem}
\section{Diagonalization}
\begin{mdframed}
  A square matrix A is said to be diagonalizable if, there exists an invertible matrix $P$ and a diagonal matrix $D$, such that $$A = PDP^{-1}.$$ If $A = PDP^{-1}$ then $A$ is said to be similar to $D$. \\
  This is standardize as if, $A$ is similar to a diagonal matrix.
\end{mdframed}
\vspace{4mm}

If $A = PDP^{-1}$ then prove that $A^k = PD^kP^{-1}$.
\clearpage

\subsection{The Diagonalization Procedure}
Suppose $A$ is an $n \times n$ matrix with $n$ different eigenvalues: $\lambda_1, \lambda_2, ..., \lambda_n$ so that the set of $n$ corresponding eigenvectors $v_1, v_2, ..., v_n$ are linearly independent. Now, $Av_1=\lambda _1 v_1, Av_2=\lambda _2 v_2, ...., Av_n = \lambda _n v_n$. Let $P$ be the matrix whose columns are the eigenvectors, i.e $P = [v_1, \; v_2, \; ... \; v_n] $. Then
\begin{equation}
  \label{eq:1}
AP = [\lambda _1 v_1, \; \lambda_2 v_2, \; ... \; \lambda_n v_n] =
 [v_1, \; v_2, \; , ..., \; v_n] \begin{bmatrix}
  \lambda_1 & 0 & ... & 0 \\
  0 & \lambda_2 & ... & 0 \\
  ... \\
  0 & 0 & ... & \lambda_1 \\
\end{bmatrix} = PD
\end{equation}
where $ D = \begin{bmatrix}
  \lambda_1 & 0 & ... & 0 \\
  0 & \lambda_2 & ... & 0 \\
  ... \\
  0 & 0 & ... & \lambda_1 \\
\end{bmatrix} $, so that $AP = PD$ gives, $A = P D P^{-1}$. This is summarized as the following theorem.

@@

\subsection{Exercise}
Diagonalizable the following matrix, if possible.
$$A=
\begin{bmatrix}
  5 & 0 & 0 & 0 \\
  0 & 5 & 0 & 0 \\
  1 & 4 & -3 & 0 \\
  -1 & -2 & 0 & -3
\end{bmatrix}
$$

\subsection{Similarity}
If $A$ an $B$ are $n \times n$ matrices, then $A$ \textbf{similar to} $B$ if there is an invertible matrix $P$ such that $$A = PBP^{-1}$$ or equivalently $B=P^{-1}AP$. Writing $Q=P^{-1}$, we can say that $B$ is similar to $A$, and we simply say $A$ and $B$ are similar.
\begin{theorem}
  If $n \times n$ matrices $A$ and $B$ are similar, then they have the same characteristic polynomial and hence the same eigenvalues (with the same multiplicities).
\end{theorem}
\begin{proof}
 \hspace{1cm} Given, $B=P^{-1}AP$ we have, $(B-\lambda I) = P^{-1}AP - \lambda P^{-1}P = P^{-1}(A- \lambda I) P$. Then
    \begin{align*}
      det(B-\lambda I) &= \det[P^{-1}(A- \lambda I) P] \\
                       &= \det(P^{-1}).\, \det(A- \lambda I).\, \det(P) \\
                       &= \det(P^{-1}).\, \det(P). \, \det(A- \lambda I) \\
                       &= \det(A- \lambda I)
    \end{align*}
  As, $\det(P^{-1}).\, \det(P) = \det(P^{-1}.P)=det(I) = 1$. Hence $A$ and $B$ has same characteristic polynomial. And the proof follows.
\end{proof}
\end{document}
